Tree establishment in models of forest dynamics: a key priority for further research

### 2023-02-24 manuscript/manuscriptRW.Rmd
revision until research recomendations

In spite of this criticism, it is remarkable that most models were not
overly off with respect to simulated ingrowth levels -- after all, few
if any of the models' establishment routines are sufficiently
constrained by data, with the exception of the empirical models xComp
and SIBYLA. It is noteworthy that in spite of their empirical basis,
both models were off relative to empirical ingrowth levels, in similar
magnitude as the other models and even in opposite directions (SIBYLA:
underestimation; xComp: overestimation). It appears that using empirical
data of limited geographical scope, as done in either of these models,
leads to extrapolation problems already under current climatic
conditions. For the other models, whose establishment module is not
strongly constrained by empirical data, the degrees of freedom that are
available for modeling ingrowth [@konig_tree_2022] could have implied
that model performance would be much worse.

Ingrowth levels define the structure and composition of future forests,
but it is equally important to correctly identify areas where ingrowth
is lacking [@rammer2021]. There are multiple constraints to the
regeneration niche of tree species [@price2001], and therefore the
absence of ingrowth is likely to be common [@fortin2007], even over
larger areas such as the one-hectare plots used here. Tree ingrowth data
are characterized by zero-inflation. This was clearly evident from the
EuFoRIa dataset, but some of the models did not produce zeros at all, or
featured a very low proportion of zero data. This substantial difference
may be due to the fact that the simulation results were drawn from
equilibrium forests, whereas in reality many of the forest reserves are
recovering from past management activities and have become denser and
darker over the past decades [e.g., @heiri2009], leading to less
ingrowth than in an equilibrium situation.

Competition for light as a strong filter for tree establishment has been
documented widely [@messier1999; @collet2006; @berdanier2016]. However,
it is difficult to measure light availability at large spatial and
temporal scales. We used total stand basal area as a proxy for light
availability [cf. @schmid2021]. We found pronounced differences in the
stand basal area ranges simulated by the models, but also between the
models and the observations. This made it impossible to evaluate the
establishment for the extremes of the density ranges in some models. For
example, ingrowth rates at low stand densities are relevant to assess
how well forests are recovering e.g., after gap creation due to
disturbance [@seidl2022; @grubb1977]. At the level of the simulated
one-hectare samples, average basal area is typically not very low as
long as no larger disturbances occur, which was explicitly excluded in
our simulation protocol. Towards the other end of the spectrum, i.e.,
with increasing basal area, it would be reasonable to expect that with
increasing stand basal area, the ingrowth of the different species would
become sparser and drop out entirely under low-light conditions
[@klopcic2012; @zell2019]. However, few models showed this trend, thus
indicating that the relation between ingrowth and light availability is
not yet captured correctly in most models.

@@

Correctly capturing species composition in tree establishment is
important to assess the future functional diversity of a forest, e.g.,
its sensitivity to drought or resilience to disturbances [@redmond2015;
@seidl2022]. In the simulations, overall ingrowth species diversity
levels were well within the observed range for two thirds of the models
(10 out of 15). Thus, while most of the models are facing difficulties
to quantitatively match ingrowth levels (cf. above), their performance
is much better when it comes to matching the quality, i.e., the
composition of ingrowth as a function of abiotic and biotic conditions.

Yet, there were pronounced variations how the main tree species were
represented along the environmental gradients. Most models maintained or
even increased ingrowth diversity between the 7 and 10 cm thresholds,
and the same is visible from the empirical data, but the differences are
not significant. Diversity variations may be explained by the way the
individual models are considering establishment processes
[@konig_tree_2022], e.g., in the absence of adults, ingrowth can be
linked to either seed rain-dispersal, or to a seed/seedling bank. The
ingrowth rates were sampled from simulations in the equilibrium, and in
this state it is expected that non-climax species such as *Pinus
sylvestris* or *Quercus* would be of minor importance, or absent
entirely [@klopcic2015]. Most models captured this difference, which is
also found in the empirical data. Thus, although the broad patterns are
matched by many models, improvements in the quantification of the
regeneration niche of the species are needed, but this cannot be done in
the absence of robust datasets.

The species diversity of the entire stand was better captured by the
models than the diversity of tree ingrowth, and only a few models
overpredicted stand-level diversity. Defining ingrowth composition as
being proportional to adult abundance, regardless of productivity, might
be a simple and conservative assumption to safely maintain relative
species abundances [@hanbury-brown2022], although this may be
simplistic. Furthermore, based on the simulation results there is no
evidence that models with feedback captured better the ingrowth species
diversity than those without feedback. The similar performance of models
with and without feedback is likely because the models are giving more
weight to the regeneration niche arising abiotic and biotic filters than
from the habitat niche of the adult trees [@grubb1977].

@@

There are several factors that lead to mortality in establishment such
as competition [@casper1997] in addition to multiple abiotic factors
[@schmid2021; @cunningham2006]. As mentioned above, it is reasonable to
expect that models that overestimate ingrowth of small trees will have a
particularly high mortality between the 7 and 10 cm thresholds. However,
this was not consistently evident from the simulations. This implies
that these models must have the excess of mortality in earlier or later
stages, if we assume that all models would be able to capture the
structure and composition of adult stands along the EuFoRIa gradient,
which however was not tested here.

Higher mortality towards the adult stage has important implications for
forest dynamics and the goods and services provided by forests. On the
one hand, mortality in later stages may erroneously enhance the share of
less shade tolerant species in the models [@klopcic2015], therefore
shifting the species composition. Unrealistic high stem densities for a
longer period of time may overestimate the role of tree regrowth in
carbon sink dynamics [@pugh2019]. Given our set of simulations and
analyses, we cannot conclusively assess what is going on in the models,
and further studies are required that focus on a wider range of tree
sizes and the fate of tree establishment along such a size continuum.

@@

The EuFoRIa data as used in this study is unique and highly suited to
better understand tree establishment. However, three aspects of this
data may also represent considerable limitations. First, we made a
comparison of tree ingrowth in a posited equilibrium state, but we
cannot assess how close the empirical data were to such a state. The
data were collected in forest reserves where no management has taken
place for long periods of time. This makes our assumption of an
equilibrium between forest properties and environmental drivers less
haphazard than it might appear at first sight. However, it may explain
why simulated ingrowth rates were higher than the observed ones. Second,
the data were collected for ingrowth above a 7 cm threshold, thus
limiting the assessment of tree establishment in a holistic sense. This
constituted a hard limit based on which we can understand some aspects
of tree establishment, which in its entirety often comprises a rather
long period since germination [@price2001]. However, many environmental
constraints are acting on young trees [@käber2021] that we were unable
to assess. Yet, ingrowth data with lower thresholds are simply not
available in a harmonized manner across large environmental gradients.
Third, the empirical data were collected from rather small plots, while
we sampled simulated ingrowth rates from 1 ha areas, which may lead to
an incorrect representation of space. Even though this strategy it is
not ideal, it represents a common challenge when harmonizing diverse
data sources originating from varied sampling strategies [@portier2022].
It would have been extremely challenging for such a variety of models to
follow a protocol where the spatial sampling size was different at each
of the 200 sites.

The design of our sampling protocol did not include spatial aspects. We
considered a wide range of models of forest dynamics, from stand to
global scale. However, the simulation setup was limited to 1 ha ingrowth
values. The lack of spatial considerations potentially puts landscape
models at a disadvantage, as they have been built to be accurate at the
landscape level. Without spatial context, we are limiting ingrowth to
the seed influx from the stand itself, if the model does not have a
background seed input. However, Global Dynamic Vegetation Models should
not be in disadvantage due to the limited spatial considerations [but
cf. Snell et al. -@snell2014] as they are usually lacking dispersal
between cells as they are based on an abstraction of horizontal space,
or even the trees themselves [@hanbury-brown2022].

@@

Increasing complexity in establishment modules has been motivated by
better process understanding and enhancing of model accuracy
[@bugmann2022; @bugmann2019]. However, more complex models do not
necessary lead to better projections and rather to a reduced
transparency and predictive power [@franklin2020]. The results from this
study showed that increasing complexity in the establishment modules did
not represent an improvement in the projections of ingrowth levels,
species composition or mortality at early tree stages. Thus, the
question what level of detail one has to use to model establishment
processes remains open and must be addressed in future research
[@konig_tree_2022; @bugmann2022].

If it should be adequate to include more detail in the establishment
models, this will come with higher parameterisation efforts, which will
most likely lead to lower generalization as the required data will have
to be collected from specific locations, as there is no general
establishment dataset available. We recommend on the one hand that the
further development of the establishment modules is implemented as
additional features that can be traced back, as done in this study for
the model variants of ForClim and ForCEEPS, and that model complexity
and structure must always be connected with model objectives
[@albrich2020]. On the other hand, we emphasize that datasets like
EuFoRIa used here are invaluable and should be further expanded in both
their spatial extent (e.g., towards boreal and Mediterranean conditions)
as well as in time (e.g., continuing the monitoring into the future).

Based on the results of this study, it is clear that models of forest
dynamics need a focus on their establishment modules. However, here we
have only consider tree establishment in the equilibrium state. It is
equally important to understand how these models project tree
establishment after changes in forest structure like those brought about
by disturbances [@seidl2022], or under different management strategies
[@lindner2000]. However, this will require an entirely different set of
observed data, and potentially not all models of forest dynamics would
be able to assess the relationship of these aspects on tree
establishment, e.g., due to the lack of disturbance or appropiate
management modules.

>>>>

## Model formulation did not affect model performance

ADD THESE POINTS HERE complexity, empirical v. PB, etc.

In spite of this criticism, it is remarkable that most models did not
exceedingly deviate from respect to simulated ingrowth levels -- after
all, few if any of the models' establishment routines are sufficiently
constrained by data, with the exception of the empirical models xComp
and SIBYLA. It is noteworthy that in spite of their empirical basis,
both models did not match empirical ingrowth levels, in similar
magnitude as the other models and even in opposite directions (SIBYLA:
underestimation; xComp: overestimation). It appears that using empirical
data of limited geographical scope and constrained to managed forests,
as done in either of these models, leads to extrapolation problems
already under current climatic conditions. For the other models, whose
establishment module is not strongly constrained by empirical data, the
degrees of freedom that are available for modeling ingrowth
[@konig_tree_2022] could have implied that model performance would be
much worse.

Ingrowth levels define the structure and composition of future forests,
but it is equally important to correctly identify areas where ingrowth
is lacking [@rammer2021]. There are multiple constraints to the
regeneration niche of tree species [@price2001], and therefore the
absence of ingrowth is likely to be common [@fortin2007], even over
larger areas such as the one-hectare plots used here. Tree ingrowth data
are characterized by zero-inflation. This was clearly evident from the
EuFoRIa dataset, but some of the models did not produce zeros at all, or
featured a very low proportion of zero data. This substantial difference
may be due to the fact that the simulation results were drawn from
equilibrium forests, whereas in reality many of the forest reserves are
recovering from past management activities and have become denser over
the past decades [e.g., @heiri2009], leading to less ingrowth than in an
equilibrium situation.

Competition for light as a strong filter for tree establishment has been
widely documented [@messier1999; @collet2006; @berdanier2016], but
models did not reproduce this expectation. However, it is difficult to
measure light availability at large spatial and temporal scales. We used
total stand basal area as a proxy for light availability [cf.
@schmid2021]. We found pronounced differences in the stand basal area
ranges simulated by the models, but also between the models and the
observations. This made not possible to evaluate the establishment for
the extremes of the stand density ranges in some models. For example,
ingrowth rates at low stand densities are relevant to assess how well
forests are recovering e.g., after gap creation due to disturbance
[@seidl2022; @grubb1977]. At the level of the simulated one-hectare
samples, average basal area is typically not very low as long as no
larger disturbances occur, which was explicitly excluded in our
simulation protocol. Towards the other end of the spectrum, i.e., with
increasing basal area, it would be reasonable to expect that with
increasing stand basal area, the ingrowth of the different species would
become sparser and drop out entirely under low-light conditions
[@klopcic2012; @zell2019]. However, few models showed this trend, thus
indicating that the relation between ingrowth and light availability is
not yet captured correctly in most models.

@@

Correctly capturing species composition in tree establishment is
important to assess the future functional diversity of a forest, e.g.,
its sensitivity to drought or resilience to disturbances [@redmond2015;
@seidl2022]. In the simulations, overall ingrowth species diversity
levels were well within the observed range for two thirds of the models
(10 out of 15). Thus, while most of the models are facing difficulties
to quantitatively match ingrowth levels (cf. above), their performance
is much better when it comes to matching the composition of ingrowth as
a function of abiotic and biotic conditions. Most models maintained or
even increased ingrowth diversity between the 7 and 10 cm thresholds,
and the same is visible from the empirical data, but the differences are
not significant. Diversity variations may be explained by the way the
individual models consider establishment processes [@konig_tree_2022],
e.g., in the absence of adults, ingrowth can be linked to either seed
rain-dispersal, or to a seed/seedling bank.

Yet, there were pronounced variations how the main tree species were
represented along the environmental gradients, in particular the
dominant species Abies alba, Fagus sylvatica and Picea abies. The
ingrowth rates were sampled from simulations in the equilibrium, and in
this state it is expected that non-climax species such as *Pinus
sylvestris* or *Quercus* would be of minor importance, or absent
[@klopcic2015]. Most models captured this low abundance, which is also
found in the empirical data. Thus, although the broad patterns are
matched by many models, improvements in the quantification of the
regeneration niche of the species are needed, but this cannot be done in
the absence of robust datasets.

The species diversity of the entire stand was better captured by the
models than the diversity of tree ingrowth, and only a few models
overpredicted stand-level diversity. Defining ingrowth composition as
being proportional to adult abundance, regardless of productivity, might
be a simple and conservative assumption to safely maintain relative
species abundances [@hanbury-brown2022], although this may be
simplistic. Furthermore, based on the simulation results, there is no
evidence that models with feedback captured better the ingrowth species
diversity than those without feedback. The similar performance of models
with and without feedback is likely because the models put more weight
to the regeneration niche arising from abiotic and biotic filters, than
from the habitat niche of the adult trees [@grubb1977].

@@

There are several factors that lead to mortality during the
establishment phase such as competition [@casper1997] and multiple
abiotic factors [@schmid2021; @cunningham2006]. As mentioned above, it
is reasonable to expect that models that overestimate ingrowth of small
trees will already have a particularly high mortality between the 7 and
10 cm thresholds. However, this was not consistently evident from the
simulations. This implies that these models must have the excess of
mortality in earlier or later stages, if we assume that all models would
be able to capture the structure and composition of adult stands along
the EuFoRIa gradient, which however was not directly tested here.

Higher mortality towards the adult phase has important implications for
forest dynamics and the goods and services provided by forests. On the
one hand, mortality in later stages may erroneously enhance the share of
less shade tolerant species in the models [@klopcic2015], if available,
therefore shifting the species composition. Unrealistic high stem
densities for a longer period of time may overestimate the role of tree
regrowth in carbon sink dynamics [@pugh2019]. Given our set of
simulations and analyses, we cannot conclusively assess what is
happening in the models, and further studies are required that focus on
a wider range of tree sizes and the fate of tree establishment along
such a size continuum.

@@

The EuFoRIa data as used in this study are unique and well suited to
better understand tree establishment. However, three aspects of these
data may also represent considerable limitations. First, we made a
comparison of tree ingrowth in a equilibrium state, but we cannot assess
how close the forests included in the EuFoRIa dataset were to such a
state. The data were collected in forest reserves where no management
has taken place for long periods of time. This makes our assumption of
an equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. However, the equilibrium
assumption may explain why simulated ingrowth rates were higher than the
observed ones. Second, the data were collected for ingrowth above a 7 cm
threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand some aspects of tree establishment, which in its entirety
often comprises a rather long period since seed production [@price2001].
However, many environmental constraints are acting on young trees
[@käber2021] that we were unable to assess. Yet, ingrowth data with
lower thresholds are simply not available in a harmonized manner across
large environmental gradients. Third, the empirical data were collected
from rather small plots, while we sampled simulated ingrowth rates from
1 ha areas, which may lead to an incorrect representation of space. Even
though this strategy is not ideal, it represents a common challenge when
harmonizing diverse data sources originating from varied sampling
strategies [@portier2022]. It would have been extremely challenging for
such a variety of models to follow a protocol where the spatial sampling
size was different at each of the 200 sites and it would have introduce
additional uncertainty into the results.

The design of our sampling protocol did not include spatial aspects such
as seed migration nor detailed soil data. We considered a wide range of
models of forest dynamics, from stand to global scale. However, the
simulation setup was limited to 1 ha ingrowth values. The lack of
spatial considerations potentially puts landscape models at a
disadvantage, as they have been built to be accurate at the landscape
level. Without spatial context, we are limiting ingrowth to the seed
influx from the stand itself, if the model does not have a background
seed input. However, Global Dynamic Vegetation Models should not be in
disadvantage due to the limited spatial considerations [but cf. Snell et
al. -@snell2014] as they are usually lacking dispersal between cells as
they are based on an abstraction of horizontal space, or even the trees
themselves [@hanbury-brown2022]. Detailed data on soil conditions were
not available from the observed data and indepentdent data for soil and
climatic water balance were used instead. Several models, however, used
more detailed drought stress indicators based e.g., on soil water
holding capacity that had to be derived from a rough soil quality
measure which could explain the poor performance of the models against
climatic water balance.

@@

Based on the results of this study, it is clear that models of forest
dynamics need a focus on their establishment modules. Increasing
complexity in establishment modules has been motivated by better process
understanding and enhancing of model accuracy [@bugmann2022;
@bugmann2019]. However, more complex models do not necessarily lead to
better projections and rather to a reduced transparency and predictive
power [@franklin2020]. The results from this study showed that
increasing complexity in the establishment modules did not represent an
improvement in the projections of ingrowth levels, species composition
or mortality at early tree stages. Thus, the question what level of
detail one has to use to model establishment processes remains open and
must be addressed in future research [@konig_tree_2022; @bugmann2022].

We recommend that the further development of the establishment modules
is implemented as additional features that can be traced back, as done
in this study for the model variants of ForClim and ForCEEPS, and that
model complexity and structure must always be connected with model
objectives [@albrich2020]. If it should be necessary to include more
detail in the establishment models, this will come with higher
parameterisation efforts, which will most likely lead to lower
generalization as the required data will have to be collected from
specific locations, as there is no general establishment dataset
available.

More effort should be invested into collecting harmonised datasets on
tree establishment. We emphasize that datasets like EuFoRIa used here
are invaluable and should be further expanded in both their spatial
extent (e.g., towards boreal and Mediterranean conditions) as well as in
time (e.g., continuing the monitoring into the future). Such a data will
allow a better evaluation of forest models and help to reduce the
uncertainty in their projections, which is crucial when they are used as
tools for predicting potential impacts of climate change.

Here we have only consider tree establishment in the equilibrium state.
It is equally important to understand how these models project tree
establishment after changes in forest structure like those brought about
by disturbances [@seidl2022], or under different management strategies
[@lindner2000]. However, this will require an entirely different set of
observed data, and potentially not all models of forest dynamics would
be able to assess the relationship of these aspects on tree
establishment, e.g., due to the lack of disturbance or appropriate
management modules.
